Moisture effects on leaf litter are species-specific and result in non-additive flammability in mixed conifer forest

### 2019-07-25 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Normalize and spell check

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{ Moisture content is a strong determinant of forest fuel flammability
  and varies on small spatial scales and over short time periods. Emperical
  modeling of equilibrium dead fine fuel moisture is well established, but less
  is known about the mechanisms by which species-specific litter traits
  influence litter moisture dynamics, particularly the processes of moisture
  absorption and desorption. Characterizing how species’ leaf litter retain and
  release moisture through time is critical to the assessment of fuel’s
  availability to burn. Furthermore, Natural forests are often comprised of
  mixed stands, leading to mixtures of leaf litter on the surface fuelling the
  fires. Multiple studies have established the existence of non-additive
  effects in the flammability of dry litter, but we lack an understanding of
  moisture synergies in mixed litter and of moisture effects on the
  flammability of mixed litter. The litter from eight species of a
  mixed-conifer forest was saturated and allowed to dry to determine moisture
  absorption capacity and desorption rates. Burn trials were performed in moist
  litter beds of single species and of mixtures to establish flammability
  response to the dry-down process. Moisture dynamics vary across species with
  higher specific leaf area being associated with higher maximum absorption and
  lower litter bulk density being associated with faster drying rates. Species
  that produce more aerated litter beds (such as \emph{Q. kelloggii} and
  \emph{Pinus} species) have faster rates of litter drying. We examined two
  axes of flammability variation (spread rate and fuel consumption) and found that these
  differing moisture dynamics result in time since wetting having strongly
  differing effects on flammability across the major litter types typified by
  different tree genera. Finally, we found that litter mixtures exhibited
  consistent non additivity in flame spread rate in fuel consumption. This non
  additivity became more positive with fuel drying. Litter mixtures burn with
  fire behavior more similar to that of the most flammable constituent species
  and this effect increases as fuels dry. ADD Synthesis statement.}

@@

Fire scientists often categorize fuel by size classes that correspond to
moisture exchange rates, although such simplified categories have long been
recognized as not sufficient to describe species-specific effects on moisture
dynamics \cite{Anderson-1985}. Fuel moisture dynamics of leaf litter are
dependent upon forest structure (shading and rain throughfall effects), litter
physical structure, and chemical composition of the fuelbeds
\cite{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
models describing moisture dynamics of fine fuels are used by fire scientists
\cite{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
usually aim to describe equilibrium conditions and at least include temperature
and relative humidity although some include precipitation and solar radiation
effects as well. Explicit models of fire behavior as a function of time since
wetting could improve fire prediction, however. In many litter fuel driven
systems, natural lightning ignitions are concurrent with precipitation and,
therefore, fire behavior can depend on rate of drying following a wetting
event. If species differ in their drying rate, this may strongly influence
relatively flammability across species at any particular time following wetting.
Such differential moisture content across time, may contribute to the often
high spatial heterogeneity of surface fires \cite{Knapp_Schwilk_etal-2006,
  Kreye_Hiers_etal-2018}.

@@

Litter packing (measured as litterbed density or packing ratio) controls the
flammability of dry litter fuels in seasonally dry climates because higher
litter density reduces flammability, especially on the flame spread rate axis.
Leaf litter density is driven by leaf size and shape
\cite{Fonda+Belanger+etal-1998, Scarff+Westoby-2006, Kane+Varner+etal-2008}.
These litter particle effects on litter bed aeration, in turn, have effects on
litter bed flammability \cite{Ganteaume+Marielle+etal-2011,
  Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. These species-specific effects
are well established, however, how moisture dynamics modulate species specific
flammability is poorly known, and could be driven by by leaf traits, emergent
litter litter traits, or both. Species specific effects on litter flammability
are recorded for the forests of the southeastern US \cite{Nowacki+Abrams-2008}
where species-specific litter drying rates play a role in flammability
feedbacks \cite{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of
the western USA, where the fire regime is characterized by surface fires fueled
by plant litter, overstory tree composition drives variation in local fire
behavior \cite{Schwilk+Caprio-2011} as a result of differential litter
flammability cross tree species and as a result of synergistic interactions in
multi-species litter mixtures \cite{Magalhaes+Schwilk-2012}. These flammability
feedbacks may help explain historical fire frequency patterns, but current
estimates of litter flammability in mixed conifer forests are based largely on
burning dry litter.

Fuel moisture content through time is controlled by two main processes, absorptive capacity and rate of moisture loss \citep{Kreye+Kobziar+etal-2013}. Absorptive capacity represents the potential maximum fuel moisture and is evaluated by the moisture content of litter when saturated. After saturation, the differences in moisture content indicate different abilities to retain moisture. Litter particle geometry may have direct effects on maximum moisture absorption through area-mass relationships and we also expect effects of leaf morphaology, espcially leaf cuticle thickness \cite{Van_Wagner-1969}. Drying rates should also be influenced by leaf and litter particle traits: surface-area-to-volume ratio and litter bulk density will influence air flow through the litter bed. Litter packing therefore has two pathways of effect on fire behavior: directly by its effect on oxygen availability for combustion, and indirectly by its effect on dry-down rate, and both these pathways are affected by leaf litter traits, especially leaf size \cite{Scarff+Westoby-2006}. Fuelbeds composed of larger leaves with lower moisture retainability and should favor drying, thus increasing the probability of ignition and fire spread.

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In litter driven fire, the two major
axes of variation are described by measurements related to flame spread rate
and measurements related to heat transfer. Flame spread rate is easy to measure
in burning trials, but heat transfer can be more difficult and often proxies
such as duration of heating are used \citep{Magalhaes+Schwilk-2012,
  Varner+Kane+etal-2015}. These different components of fire behavior have
differing ecological effects. Species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such non-additive effects have been shown to be synergistic in dry
litter fuels where the flammability of a mixture is driven by the most
flammable species \cite{VanAltena+Logtestjin+etal-2012,
  Magalhaes+Schwilk-2012}. It is possible that similar non-additive effects may
influence litter moisture content, especially if different leaf traits
influence adsorptive capacity and drying rate. There is some evidence for
non-additive flammability in moist fuels with one study finding increasing
non-additivity with moisture \cite{Blauw+Wensink+etal-2015}.

@@

% The dry-down curves indicated that the eight species varied in their moisture behavior (Fig \ref{fig-drydown}). To determine a subset of species to use in the mixture trials, we performed analysis of contrasts using lsmeans package in R (TODO Lenth 2016) to guide choosing a subset of species for the litter mixtures. Based on the marginal means analysis, we chose four species to maximize variation in water retention and drying rate. Creating all possible three species mixtures from these four species resulted in four distinct three-species mixtures.

% Leaf trait data

@@

We performed flammability tests across multiple moisture levels on monospecific
litter beds and on mixed litter beds with each species or mixture replicated 5
times at each moisture level. We used the initial dry-down experiment to guide
our timeing so as to attempt to burn samples over a range of moisture contents.
We aimed for moisture contents from 10\% to 100\% on a dry mass basis as
preliminary trials on damp fuels suggested that across these fuels ignition is
unlikely above 100\% fuel moisture. To expand our data on the dry end of this
gradient, we supplemented these data with data from an earlier experiment on
oven-dried litter \cite{Magalhaes+Schwilk-2012}. This supplemental data added
five replicates per species at an oven dried moisture content of $<$ 5\%. We
assigned these replicates a time since wetting value of 144 hours equivalent to
the longest time since drying for the wetted samples. Flammability was assessed
using a 150\,cm long burn table, 15\,cm wide and 15\,cm tall, in which leaf
litter was placed and the table gently shaken to allow settling. Burning trials
all used 450\,g (dry weight) of litter. For three-species mixtures, this was
150\,g per species. This design allows fire to reach constant flame spread
rates \cite{Magalhaes+Schwilk-2012}. Two graduated metal rulers equally spaced
along the apparatus allowed visual determination of maximum flame height.
Litter was ignited with a propane torch; maximum adiabatic flame temperature of
1899 C, \url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the
apparatus and allowed to burn until extinction. If a sample failed to ignite
after 30 seconds, no further attempt was made and if three samples of a species
at a particular mositure level failed to ignite, no further attempts were made
on that species/moisture combination and the replicate was allowed to continue
drying. A timer was used to record total time to ignition and duration of
flaming combustion. The flammability trials followed the methodology detailed
in de Magalhães and Schwilk \cite{Magalhaes+Schwilk-2012}.

% and nine type-K thermocouples attached to data loggers measured temperature every second at three distances from the ignition point and at three different depths below the litter surface.

@@

Where $m$ is moisture content at time $t$, and $\lambda$ is the exponential
decay coefficient. To estimate these coefficents we fit a linear mixed-effects
model by first taking the natural logarithm of moisture content, with time
since wetting and species as fixed effects nested within replicate basket. We
extracted the coefficients of the fitted model to describe maximum water
retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate
($\lambda$; hour$^{-1}$) for each species. All analyses were conducted in R
\cite{RCoreTeam-2019}. We fit models with R packages `lme4'
\cite{Bates_Machler_etal-2015} and `afex' \cite{Singmann_Bolker_etal-2017}.
Approximate degrees of freedom and p-values were calculated by the
Kenward-Roger approximation \cite{Kenward_Roger-1997} which is recommended by
Luke \cite{Luke-2017} as producing acceptable Type 1 error rates. We tested if
one leaf trait (specific leaf area) and one litter trait (litter bulk density)
influenced moisture dynamics by fitting linear models with species means as
observations. For each response variable we fit a single linear model with two
predictors and an interaction term.

@@

We investigated possible non-additive effects on moisture and on flammability
by testing if moisture content and flammability of mixtures were predicted by
the average of the constituent species. To choose mixtures, we first grouped
the species according to their maximum water retainability and dessication rate
according to estimated marginal means by species using the R package `emmeans'
\cite{Lenth-2019}. This resulted in four groups of species and we then selected
one representative species from each litter type group and produced all four
possible three-species litter mixtures using those four species. For each
mixture and time since wetting, we predicted moisture content based on the mean
content of the three consituent species at that time since wetting. We
predicted flammability parameters based on the average flammability measures
from individual species trials at each time since wetting and then calculated a
predicted spread rate and fuel consumption as the average of these three values
for the mixture. The expectation from the null model was that the difference
between observed and predicted values of flammability will be zero. We tested
if there were significant departures from the null (zero) with a mixed effects
linear model. For flammability parameters, we fit linear mixed effect models
with the observed-predicted flammability value as the dependent variable and
used Wald tests to test if the centered intercept was different than zero (a
measure of overall non-additivity) and if the slope with time since wetting was
different from zero (indicating non-additivity changing with drying).

@@

There were marked differences in the extent of moisture species retain initially
following wetting and draining by gravity (all $P$ < 0.0001, Fig.
\ref{fig:drydown} and Table \ref{tab:drydown}). \emph{Q. kelloggii}, the only
broadleaf, is the species that retained the most moisture initially.
\emph{Quercus}, like the pines, had a rapid drying rate (Fig.
\ref{fig:drydown} and SI Fig. 1). Calculating estimated marginal means following the
linear mixed-effects model discriminated the species into three different
maximum water retention groups: 1) \emph{Q. kelloggii}; 2) \emph{P.
  lambertiana} and \emph{C. decurrens}; and 3) \emph{A. magnifica}, \emph{S.
  giganteum}, \emph{A. concolor}, \emph{P. jeffreyi} and \emph{P. ponderosa}
(SI Fig. 1). There are also differences in desiccation rates across the eight
species and estimated mixed linear model slopes discriminated species into
three groups from lowest to most rapid: 1) \emph{C.decurrens}, 2) \emph{Abies
  spp} and \emph{S. gianteum}, and 3) \emph{Q. kelloggii} and \emph{Pinus spp}
(SI Fig. 2).

@@

\begin{table}[H]
  \caption{Generalized linear mixed model results for two sets of models predicting fuel consumption. The first set modeled fuel consumption as a function of fuel moisture (mc) with taxonomic group as a fixed effect and species as a nested random effect. The second set modeled fuel consumption as a function of as a function of time since wetting (time) with taxonomic group as a fixed effet and species as a nested random effect. The analysis of deviance table below shows sequential chi-square tests on nested beta regression models: from the simplest null model within each group with only intercept and random effect, for a taxon effect only model, for a taxon and moisture content effect, and for the full model including the taxon by moisture or time interaction. Coefficient estimates are in SI Tables 3 and 5.}
  \label{tab:consume_anova}
  \centering

@@

As a consequence of the relationships shown above, time since wetting had a significant effect on spread rate ($P$ < 0.0001) and the slope of this effect varied by taxonomic group (Fig. \ref{fig:spread_time}, Table \ref{tab:spreadrate_anova}). The pines were able to ignite and burn much sooner after wetting than did the other taxa, none of which had measurable spread rates until burning was attempted 72 hours after wetting (note that trials that failed to even ignite smoldering combustion are ommitted completely from the model and figure). At that point, the \emph{Abies}, \emph{Calocedrus}, and \emph{Sequiadendron} had dryer litter than did the \emph{Quercus}, yet the spread rates were slow. \emph{Quercus} had the steepest response to time since wetting (Fig. \ref{fig:spread_time}, SI Table 4)).

@@

Fuel consumption was highly variable across trials and intermediate values were less common than were very low or moderately high consumption (Fig. \ref{fig:consume_time}). All parameters significantly contributed to the generalized linear mixed beta regression model as indicated by the nested model comparisons in Table \ref{tab:consume_time_anova}. As with flame spread rate some pines had the most rapid response to drying. However, unlike in the other flammability analyses, here the pines did not all behave similarly: \emph{P. lambertiana} had lower fuel consumption than did \emph{P. ponderosa} and \emph{P. jeffreyii}. We noticed this during trials and in model fitting we found that placing \emph{P. lambertiana} in its own group allowed model convergence that was otherwise impossible. This was at odds with our intent to use only \emph{a priori} taxonomic groups, but we could not obtain model convergence otherwise. This result should therefore be seen as preliminary because the test does not reflect an \emph{a priori} hypothesis.

@@

Based on visual inspection of the dry down curves and the estimated marginal means for maximum water retainability and desiccation rate (SI Figs 2--3), we selected four species to represent the full range of litter moisture dynamics for use in the litter mixtures and examined all four three-species combinations of those four. The four species were \emph{Abies concolor}, \emph{Calocedrus decurrens}, \emph{Pinus jeffreyi} and \emph{Quercus kellogii}.

@@

We measured drying rates and flammability of four unique three-species litter mixtures. These mixtures contained species representative of all dry-down behavior types observed in single species trials. As expected from the single species results, mixtures that contained \emph{Q. kelloggii} absorbed more water during wetting, and the one mixture without any Cupressaceae had the most rapid drying rate (Figure \ref{fig:mixture_dry_down} and Table \ref{tab:mixtures_drydown}). Drydown rates, however, did not behave as the average of the constituent species: mixtures tended to be drier than expected early in the process and wetter than expected later (Fig. \ref{fig:mixture_obs_pred_mc}). The fact that mixtures tended to have lower water content than predicted from the mean of the constituent species after wetting, but higher water content than expected after long drying is surprising because past work has shown that
mixtures tend to pack less densely than do litter monocultures \cite[][, Table 2]{Magalhaes+Schwilk-2012}. If packing density controls rate of drying, then we would expect mixtures to dry more rapidly than predicted from monoculture trials. This pattern in mositure contrent through time, however, did not directly predict flammability through time.

The overall ranking of flammability in litter mixtures showed some consistency
with the results obtained for the individual species, but there were
significant non-additive effects. Mixtures containing \emph{Q. kelloggii} and
\emph{P. jeffreyi} tended to have fast spread rates. These results are likely
influenced by the lower bulk densities of litter mixtures containing those two
species. Most interestingly, there were striking pattern of non-additivity in
spread rate and in fuel consumption for the four mixtures, indicating that
positive non-additivity in flammability is not restricted to completely dry
litter mixtures observed in previous work \citep{Magalhaes+Schwilk-2012,
  VanAltena+Logtestjin+etal-2012}. Non-additivity in both spread rate and fuel
consumption became more positive with time since wetting and these patterns
were consistent across mixture identities. This is a slightly different result
than that reported by Blauw et al \citeyear{Blauw+Wensink+etal-2015}. Blauw et
al found that non-addivity was greater in moist mixtures of heathland fuels
than in dry fuels. In their work, non-addiitvity was generally negative but cases of 
positive non-addivity in mixtures was often a result of mixtures meeting
ignition thresholds not met by the constituent species alone. To limit the
number of models fit, we limited our investigation to two flammability
parameters that represent two major axes of flammability variation. Therefore,
our study does not fully capture ignition effects although we do have some
results consistent with theirs: mixtures with \emph{Q. kellogii} could ignite
at times since wetting at which the species in monoculture could not.

@@

Positive non-additivity, whereby more flammable species have effects on fire
behavior disproportionate to their contribution to the litter layer, could lead
to vegetation feedbacks. In this temperate mixed conifer forest, these effects
are likely to include positive feedbacks whereby the shade intolerant species
such as the long-needled pines and the large-leaved oak, are favored by the
increased fire to which their fuels contribute \cite{Schwilk+Caprio-2011} and
similar results are reported for southeastern US forests where species
composition shifts have led to decreases in fire activity
\citep{Nowacki+Abrams-2008}. Such positive feedback effects increases the
potential for species to exert community scale effects even when relatively
rare and that can influence community assembly and the evolutionary of
niche-constructing traits \cite{Kerr+Schwilk+etal-1999, Schwilk+Kerr-2002}.
Ongoing climate change and concurrent changes to fire regimes are linked
phenomena. The patterns we demonstrate here show to potential importance of species
traits and interactions at small scales and among subtle shifts in vegetation,
but vegetation-fire feedbacks are important at a wide range of scales from stand-level species composition shifts to the earth system \cite{Harris+Remenyi+etal-2016, Archibald+Lehmann+etal-2018}.

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{ Moisture content is a strong determinant of forest fuel flammability
  and varies on small spatial scales and over short time periods. Emperical
  modeling of equilibrium dead fine fuel moisture is well established, but less
  is known about the mechanisms by which species-specific litter traits
  influence litter moisture dynamics, particularly the processes of moisture
  absorption and desorption. Characterizing how species’ leaf litter retain and
  release moisture through time is critical to the assessment of fuel’s
  availability to burn. Furthermore, Natural forests are often comprised of
  mixed stands, leading to mixtures of leaf litter on the surface fuelling the
  fires. Multiple studies have established the existence of non-additive
  effects in the flammability of dry litter, but we lack an understanding of
  moisture synergies in mixed litter and of moisture effects on the
  flammability of mixed litter. The litter from eight species of a
  mixed-conifer forest was saturated and allowed to dry to determine moisture
  absorption capacity and desorption rates. Burn trials were performed in moist
  litter beds of single species and of mixtures to establish flammability
  response to the dry-down process. Moisture dynamics vary across species with
  higher specific leaf area being associated with higher maximum absorption and
  lower litter bulk density being associated with faster drying rates. Species
  that produce more aerated litter beds (such as \emph{Q. kelloggii} and
  \emph{Pinus} species) have faster rates of litter drying. We examined two
  axes of flammability variation (spread rate and fuel consumption) and found
  that these differing moisture dynamics result in time since wetting having
  strongly differing effects on flammability across the major litter types
  typified by different tree genera. Finally, we found that litter mixtures
  exhibited consistent non-additivity in flame spread rate in fuel consumption.
  This non-additivity became more positive with fuel drying. Litter mixtures
  burn with fire behavior more similar to that of the most flammable
  constituent species and this effect increases as fuels dry. ADD Synthesis
  statement.}

@@

Fire scientists often categorize fuel by size classes that correspond to
moisture exchange rates, although such simplified categories have long been
recognized as not sufficient to describe species-specific effects on moisture
dynamics \cite{Anderson-1985}. Fuel moisture dynamics of leaf litter are
dependent upon forest structure (shading and rain throughfall effects), litter
physical structure, and chemical composition of the fuel-beds
\cite{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
models describing moisture dynamics of fine fuels are used by fire scientists
\cite{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
usually aim to describe equilibrium conditions and at least include temperature
and relative humidity although some include precipitation and solar radiation
effects as well. Explicit models of fire behavior as a function of time since
wetting could improve fire prediction, however. In many litter fuel driven
systems, natural lightning ignitions are concurrent with precipitation and,
therefore, fire behavior can depend on rate of drying following a wetting
event. If species differ in their drying rate, this may strongly influence
relatively flammability across species at any particular time following
wetting. Such differential moisture content across time, may contribute to the
often high spatial heterogeneity of surface fires
\cite{Knapp_Schwilk_etal-2006, Kreye_Hiers_etal-2018}.

@@

Litter packing (measured as litter-bed density or packing ratio) controls the
flammability of dry litter fuels in seasonally dry climates because higher
litter density reduces flammability, especially on the flame spread rate axis.
Leaf litter density is driven by leaf size and shape
\cite{Fonda+Belanger+etal-1998, Scarff+Westoby-2006, Kane+Varner+etal-2008}.
These litter particle effects on litter bed aeration, in turn, have effects on
litter bed flammability \cite{Ganteaume+Marielle+etal-2011,
  Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. These species-specific effects
are well established, however, how moisture dynamics modulate species specific
flammability is poorly known, and could be driven by by leaf traits, emergent
litter litter traits, or both. Species specific effects on litter flammability
are recorded for the forests of the southeastern US \cite{Nowacki+Abrams-2008}
where species-specific litter drying rates play a role in flammability
feedbacks \cite{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of
the western USA, where the fire regime is characterized by surface fires fueled
by plant litter, overstory tree composition drives variation in local fire
behavior \cite{Schwilk+Caprio-2011} as a result of differential litter
flammability cross tree species and as a result of synergistic interactions in
multi-species litter mixtures \cite{Magalhaes+Schwilk-2012}. These flammability
feedbacks may help explain historical fire frequency patterns, but current
estimates of litter flammability in mixed conifer forests are based largely on
burning dry litter.

Fuel moisture content through time is controlled by two main processes,
absorptive capacity and rate of moisture loss \citep{Kreye+Kobziar+etal-2013}.
Absorptive capacity represents the potential maximum fuel moisture and is
evaluated by the moisture content of litter when saturated. After saturation,
the differences in moisture content indicate different abilities to retain
moisture. Litter particle geometry may have direct effects on maximum moisture
absorption through area-mass relationships and we also expect effects of leaf
morphology, especially leaf cuticle thickness \cite{Van_Wagner-1969}. Drying
rates should also be influenced by leaf and litter particle traits:
surface-area-to-volume ratio and litter bulk density will influence air flow
through the litter bed. Litter packing therefore has two pathways of effect on
fire behavior: directly by its effect on oxygen availability for combustion,
and indirectly by its effect on dry-down rate, and both these pathways are
affected by leaf litter traits, especially leaf size
\cite{Scarff+Westoby-2006}. Fuel-beds composed of larger leaves with lower
moisture retainability and should favor drying, thus increasing the probability
of ignition and fire spread.

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In litter driven fire, the two major
axes of variation are described by measurements related to flame spread rate
and measurements related to heat transfer. Flame spread rate is easy to measure
in burning trials, but heat transfer can be more difficult and often proxies
such as duration of heating are used \citep{Magalhaes+Schwilk-2012,
  Varner+Kane+etal-2015}. These different components of fire behavior have
differing ecological effects. Species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such non-additive effects have been shown to be synergistic in dry
litter fuels where the flammability of a mixture is driven by the most
flammable species \cite{VanAltena+Logtestjin+etal-2012,
  Magalhaes+Schwilk-2012}. It is possible that similar non-additive effects may
influence litter moisture content, especially if different leaf traits
influence absorptive capacity and drying rate. There is some evidence for
non-additive flammability in moist fuels with one study finding increasing
non-additivity with moisture \cite{Blauw+Wensink+etal-2015}.

@@

@@

We performed flammability tests across multiple moisture levels on monospecific
litter beds and on mixed litter beds with each species or mixture replicated 5
times at each moisture level. We used the initial dry-down experiment to guide
our timing so as to attempt to burn samples over a range of moisture contents.
We aimed for moisture contents from 10\% to 100\% on a dry mass basis as
preliminary trials on damp fuels suggested that across these fuels ignition is
unlikely above 100\% fuel moisture. To expand our data on the dry end of this
gradient, we supplemented these data with data from an earlier experiment on
oven-dried litter \cite{Magalhaes+Schwilk-2012}. This supplemental data added
five replicates per species at an oven dried moisture content of $<$ 5\%. We
assigned these replicates a time since wetting value of 144 hours equivalent to
the longest time since drying for the wetted samples. Flammability was assessed
using a 150\,cm long burn table, 15\,cm wide and 15\,cm tall, in which leaf
litter was placed and the table gently shaken to allow settling. Burning trials
all used 450\,g (dry weight) of litter. For three-species mixtures, this was
150\,g per species. This design allows fire to reach constant flame spread
rates \cite{Magalhaes+Schwilk-2012}. Two graduated metal rulers equally spaced
along the apparatus allowed visual determination of maximum flame height.
Litter was ignited with a propane torch; maximum adiabatic flame temperature of
1899 C, \url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the
apparatus and allowed to burn until extinction. If a sample failed to ignite
after 30 seconds, no further attempt was made and if three samples of a species
at a particular moisture level failed to ignite, no further attempts were made
on that species/moisture combination and the replicate was allowed to continue
drying. A timer was used to record total time to ignition and duration of
flaming combustion. The flammability trials followed the methodology detailed
in de Magalhães and Schwilk \cite{Magalhaes+Schwilk-2012}.

@@

Where $m$ is moisture content at time $t$, and $\lambda$ is the exponential
decay coefficient. To estimate these coefficients we fit a linear mixed-effects
model by first taking the natural logarithm of moisture content, with time
since wetting and species as fixed effects nested within replicate basket. We
extracted the coefficients of the fitted model to describe maximum water
retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate
($\lambda$; hour$^{-1}$) for each species. All analyses were conducted in R
\cite{RCoreTeam-2019}. We fit models with R packages `lme4'
\cite{Bates_Machler_etal-2015} and `afex' \cite{Singmann_Bolker_etal-2017}.
Approximate degrees of freedom and p-values were calculated by the
Kenward-Roger approximation \cite{Kenward_Roger-1997} which is recommended by
Luke \cite{Luke-2017} as producing acceptable Type 1 error rates. We tested if
one leaf trait (specific leaf area) and one litter trait (litter bulk density)
influenced moisture dynamics by fitting linear models with species means as
observations. For each response variable we fit a single linear model with two
predictors and an interaction term.

@@

We investigated possible non-additive effects on moisture and on flammability
by testing if moisture content and flammability of mixtures were predicted by
the average of the constituent species. To choose mixtures, we first grouped
the species according to their maximum water retainability and dessication rate
according to estimated marginal means by species using the R package `emmeans'
\cite{Lenth-2019}. This resulted in four groups of species and we then selected
one representative species from each litter type group and produced all four
possible three-species litter mixtures using those four species. For each
mixture and time since wetting, we predicted moisture content based on the mean
content of the three constituent species at that time since wetting. We
predicted flammability parameters based on the average flammability measures
from individual species trials at each time since wetting and then calculated a
predicted spread rate and fuel consumption as the average of these three values
for the mixture. The expectation from the null model was that the difference
between observed and predicted values of flammability will be zero. We tested
if there were significant departures from the null (zero) with a mixed effects
linear model. For flammability parameters, we fit linear mixed effect models
with the observed-predicted flammability value as the dependent variable and
used Wald tests to test if the centered intercept was different than zero (a
measure of overall non-additivity) and if the slope with time since wetting was
different from zero (indicating non-additivity changing with drying).

@@

There were marked differences in the extent of moisture species retain
initially following wetting and draining by gravity (all $P$ < 0.0001, Fig.
\ref{fig:drydown} and Table \ref{tab:drydown}). \emph{Q. kelloggii}, the only
broadleaf, is the species that retained the most moisture initially.
\emph{Quercus}, like the pines, had a rapid drying rate (Fig. \ref{fig:drydown}
and SI Fig. 1). Calculating estimated marginal means following the linear
mixed-effects model discriminated the species into three different maximum
water retention groups: 1) \emph{Q. kelloggii}; 2) \emph{P. lambertiana} and
\emph{C. decurrens}; and 3) \emph{A. magnifica}, \emph{S. giganteum}, \emph{A.
  concolor}, \emph{P. jeffreyi} and \emph{P. ponderosa} (SI Fig. 1). There are
also differences in desiccation rates across the eight species and estimated
mixed linear model slopes discriminated species into three groups from lowest
to most rapid: 1) \emph{C.decurrens}, 2) \emph{Abies spp} and \emph{S.
  giganteum}, and 3) \emph{Q. kelloggii} and \emph{Pinus spp} (SI Fig. 2).

@@

\begin{table}[H]
  \caption{Generalized linear mixed model results for two sets of models
    predicting fuel consumption. The first set modeled fuel consumption as a
    function of fuel moisture (mc) with taxonomic group as a fixed effect and
    species as a nested random effect. The second set modeled fuel consumption
    as a function of as a function of time since wetting (time) with taxonomic
    group as a fixed effect and species as a nested random effect. The analysis
    of deviance table below shows sequential chi-square tests on nested beta
    regression models: from the simplest null model within each group with only
    intercept and random effect, for a taxon effect only model, for a taxon and
    moisture content effect, and for the full model including the taxon by
    moisture or time interaction. Coefficient estimates are in SI Tables 3 and
    5.}
  \label{tab:consume_anova}
  \centering

@@

As a consequence of the relationships shown above, time since wetting had a
significant effect on spread rate ($P$ < 0.0001) and the slope of this effect
varied by taxonomic group (Fig. \ref{fig:spread_time}, Table
\ref{tab:spreadrate_anova}). The pines were able to ignite and burn much sooner
after wetting than did the other taxa, none of which had measurable spread
rates until burning was attempted 72 hours after wetting (note that trials that
failed to even ignite smoldering combustion are omitted completely from the
model and figure). At that point, the \emph{Abies}, \emph{Calocedrus}, and
\emph{Sequoiadendron} had dryer litter than did the \emph{Quercus}, yet the
spread rates were slow. \emph{Quercus} had the steepest response to time since
wetting (Fig. \ref{fig:spread_time}, SI Table 4)).

@@

Fuel consumption was highly variable across trials and intermediate values were
less common than were very low or moderately high consumption (Fig.
\ref{fig:consume_time}). All parameters significantly contributed to the
generalized linear mixed beta regression model as indicated by the nested model
comparisons in Table \ref{tab:consume_time_anova}. As with flame spread rate
some pines had the most rapid response to drying. However, unlike in the other
flammability analyses, here the pines did not all behave similarly: \emph{P.
  lambertiana} had lower fuel consumption than did \emph{P. ponderosa} and
\emph{P. jeffreyi}. We noticed this during trials and in model fitting we
found that placing \emph{P. lambertiana} in its own group allowed model
convergence that was otherwise impossible. This was at odds with our intent to
use only \emph{a priori} taxonomic groups, but we could not obtain model
convergence otherwise. This result should therefore be seen as preliminary
because the test does not reflect an \emph{a priori} hypothesis.

@@

Based on visual inspection of the dry down curves and the estimated marginal
means for maximum water retainability and desiccation rate (SI Figs 2--3), we
selected four species to represent the full range of litter moisture dynamics
for use in the litter mixtures and examined all four three-species combinations
of those four. The four species were \emph{Abies concolor}, \emph{Calocedrus
  decurrens}, \emph{Pinus jeffreyi} and \emph{Quercus kelloggii}.

@@

We measured drying rates and flammability of four unique three-species litter
mixtures. These mixtures contained species representative of all dry-down
behavior types observed in single species trials. As expected from the single
species results, mixtures that contained \emph{Q. kelloggii} absorbed more
water during wetting, and the one mixture without any Cupressaceae had the most
rapid drying rate (Figure \ref{fig:mixture_dry_down} and Table
\ref{tab:mixtures_drydown}). Drydown rates, however, did not behave as the
average of the constituent species: mixtures tended to be drier than expected
early in the process and wetter than expected later (Fig.
\ref{fig:mixture_obs_pred_mc}). The fact that mixtures tended to have lower
water content than predicted from the mean of the constituent species after
wetting, but higher water content than expected after long drying is surprising
because past work has shown that mixtures tend to pack less densely than do
litter monocultures \cite[][, Table 2]{Magalhaes+Schwilk-2012}. If packing
density controls rate of drying, then we would expect mixtures to dry more
rapidly than predicted from monoculture trials. This pattern in moisture
content through time, however, did not directly predict flammability through
time.

The overall ranking of flammability in litter mixtures showed some consistency
with the results obtained for the individual species, but there were
significant non-additive effects. Mixtures containing \emph{Q. kelloggii} and
\emph{P. jeffreyi} tended to have fast spread rates. These results are likely
influenced by the lower bulk densities of litter mixtures containing those two
species. Most interestingly, there were striking pattern of non-additivity in
spread rate and in fuel consumption for the four mixtures, indicating that
positive non-additivity in flammability is not restricted to completely dry
litter mixtures observed in previous work \citep{Magalhaes+Schwilk-2012,
  VanAltena+Logtestjin+etal-2012}. Non-additivity in both spread rate and fuel
consumption became more positive with time since wetting and these patterns
were consistent across mixture identities. This is a slightly different result
than that reported by Blauw et al \citeyear{Blauw+Wensink+etal-2015}. Blauw et
al found that non-additivity was greater in moist mixtures of heathland fuels
than in dry fuels. In their work, non-additivity was generally negative but cases of 
positive non-additivity in mixtures was often a result of mixtures meeting
ignition thresholds not met by the constituent species alone. To limit the
number of models fit, we limited our investigation to two flammability
parameters that represent two major axes of flammability variation. Therefore,
our study does not fully capture ignition effects although we do have some
results consistent with theirs: mixtures with \emph{Q. kellogii} could ignite
at times since wetting at which the species in monoculture could not.

@@

Positive non-additivity, whereby more flammable species have effects on fire
behavior disproportionate to their contribution to the litter layer, could lead
to vegetation feedbacks. In this temperate mixed conifer forest, these effects
are likely to include positive feedbacks whereby the shade intolerant species
such as the long-needled pines and the large-leaved oak, are favored by the
increased fire to which their fuels contribute \cite{Schwilk+Caprio-2011} and
similar results are reported for southeastern US forests where species
composition shifts have led to decreases in fire activity
\citep{Nowacki+Abrams-2008}. Such positive feedback effects increases the
potential for species to exert community scale effects even when relatively
rare and that can influence community assembly and the evolutionary of
niche-constructing traits \cite{Kerr+Schwilk+etal-1999, Schwilk+Kerr-2002}.
Ongoing climate change and concurrent changes to fire regimes are linked
phenomena. Vegetation-fire feedbacks are important at a wide range of scales
from stand-level species composition shifts to the earth system
\cite{Harris+Remenyi+etal-2016, Archibald+Lehmann+etal-2018}. The patterns we
demonstrate here show to potential importance of species traits and
interactions at local and stand scales where management and fire operations
meet species' biology.
